Closure-Free Functional Programming in a Two-Level Type Theory

### 2024-06-18 old/icfp24_old/response.txt

Review A
--------------------------------------------------------------------------------

Questions:

- What are the main formal/"soundness" results from the previous paper(s) on
  two-level type theory that section 2.4 extends?"

The main result is soundness, which means that in each program that has
an object-level type and only depends on object-level free variables, all quotes
and splices can be computed away, and the result is definitionally equal to what
we started with. This is theorem 5.2. in Kovács 22. There's also "completeness"
(roughly meaning that every object program can be represented in the staged
language) and "strictness" (meaning that the staging algorithm performs no
beta-eta conversion on object terms).

However, we glossed over completeness for CFTT (because it's easy and
mechanically follows Kovács22) and strictness holds trivially for CFTT since
there are no beta-eta rules in the object theory. Although now we think that we
should include these points in the paper revisions.

- How does the code size/complexity of your approach compare to say Haskell
  standard library definitions of monad / monad transformer operations?

For each supported transformer in the library, we need an extra Join and an
Improve instance. This is around 5 lines for each transformer in the our code
supplement. We also need let-inserting versions of operations like "put" and
"modify".

- How does performance compare on a like-for-like basis (say, with all normal
  GHC optimizations enabled and without)?
  ...
  What degree of change at the level of library/interface clients is/would be
  necessary for "normal" Haskell programs to make use of your approach?

The Template Haskell implementation is useful for TH power users but we do not
think that a) it can fully replace GHC optimizations b) ordinary users of monads
in Haskell would want to use the staged version.

For an example for a), if we write a staged "State (↑Int)" program, although the
overhead of the monadic binding will be eliminated, we still need to use GHC's
-O1 optimization to also unbox the involved "Int"-s. There are limitations on
memory layout control, data specialization and code deduplication in Haskell
which are annoying even when we use staging, and it's more convenient to just
leave some things to GHC's optimization.

For b), users would need to write explicit quotes and splices and reorganize
their code to comply with GHC stage restrictions (which means that TH code
cannot be in the same module as the splice that uses it!).

This is not to say that the TH implementation is useless; we're using it in some
of our TH code to good effect, it's just that our vision in Section 6 about
staged monads is more realistic in a standalone CFTT implementation or in a
major extension of an existing language.

Addressing other questions
------------------------------------------------------------

- discussion of ... defunctionalization ... lambda-lifting

We did cut out a discussion of defunctionalization from the paper, for lack of
space. In short, defunc by itself does not eliminate closures nor dynamic
control; a dynamic call to a heap-allocated closure becomes a dynamic switch on
a tag in a heap-allocated defunctionalized closure. In contrast, CFTT programs
compile to truly closure-free code where the only source of dynamic control flow is
case-switching on user-introduced algebraic data types. That said, if we add
closure types to CFTT, defunc can be still used in the usual way for those
closures.

As to lambda-lifting, it's actually required in CFTT compilation, for the
let-defined local functions that are not join points, i.e. called somewhere
in non-tail position.

- we get soundness of W-types

Yes, it is the soundness mentioned in our previous answer here. Perhaps more
precisely, it's "soundness of staging in the presence of W-types".

- cross-stage persistence

(This also addresses Reviewer B's related question)

CFTT in the current form does not support any CSP, like previous 2LTT-s. By this
we mean that it's not possible to write a definition that can be later reused at
both levels, without explicit conversions. For such a thing to be possible, we
would need syntax for a language fragment which can be interpreted into both
levels, perhaps as a form of "stage-polymorphism". Technically, the meta type
theory is expressive enough to embed languages that can be then interpreted into
both levels, but that would likely be less convenient to use than "native" stage
polymorphism.

We think though that the issue of code duplication across levels is mitigated in
CFTT. First, lots of code can't be duplicated because they're not even
expressible in the object language. The monads in the paper are like this: we
only have a meta-monad library. In other cases we don't really want to use the
same definitions across stages. E.g. we would use operations on push/pull
streams instead of operations on object-level lists, for most list
functions. For meta-level code we'd use ordinary non-fused lists instead,
because we expect to get sufficient performance at compile time without fancy
optimizations. Our impression of prior works in staging is that compile-time
performance tends to be good enough, and at the meta level it's better to focus on
abstraction, clarity and modularity instead of performance.

This still leaves plenty of code to be potentially duplicated,
e.g. implementations of standard immutable data structures, so there's certainly
reason to look at CSP or stage polymorphism in the future.

Thank you for the other comments as well, we'll take them into account in
revisions.

Review B
--------------------------------------------------------------------------------

Questions:

- In multiple systems with staging, only one of <~t> = t and ~<t> = t is true,
  and not always definitionally. Is this unconditionally true?

It is. It holds in the semantics in Kovács 22. Also, from an implementation
perspective, there is nothing fancy in the equational theories of 2LTT-s and
CFTT, compared to plain MLTT. Supporting ↑_ in a 2LTT is not more difficult
than supporting the following type in Agda:

  -- with eta rule
  record ↑ (A : Set₀) : Set₁ where
    constructor <_>
    field
      ~_ : A

- What about 'cross-stage persistence'?

See my answer to Reviewer A regarding CSP.

- Finally tagless for syntactic noise

The most important thing for noise-reduction is inference of quotes and
splices. It works very effectively with bidirectional elaboration, as
demonstrated in the Kovács22 supplement, mainly because universes of types carry
stage information (unlike in TH and MetaOCaml). Generally we only need to
disambiguate let-definitions, as we do in the current paper with "=" and ":=",
to be able to infer quotes and splices unambiguously. Hence, we consider
splice/quote noise to be basically solved but there may be other kinds of
staging noise in practice, like the up/down conversions.

Addressing other questions/comments
------------------------------------------------------------

- on "staging" terminology

Skimming through the literature, we're surprised and slightly horrified that
indeed, our "staging" is idiosyncratic terminology. We are open to changing
"staging" to "unstaging". Are there other choices though that we could use?

- Extra citations

We are very grateful for these pointers. The "functors and monads" and the
"combinatorial circuits" papers seem to be very relevant and we'll be sure to
mention these in the contributions section already.

As to Section 3.2, we did not intend to suggest that we invented anything
related to let-insertion and the Trick, but we do agree that we should make this
more obvious by including citations there (and not just in the related works
section). We'll also include the SOP & strymonas citations.

- "universe reflection" & "finitely presented"

In the HoTT-flavored 2LTT literature, the former is called "fibrant" and the
latter is called "cofibrant". We found these terms un-enlightening in the staged
programming context, so we skipped them, but we could mention them in a
footnote. About IsSOP, we conjecture that in CFTT it precisely captures the
notion of "finite presentation" or "cofibrancy", i.e. every cofibrant type is
isomorphic to a SOP.

- Section 2.4

This section is intended to be read as an appendix of Kovács22; it's a modest
extension but still requires a lot of technical background from there. This is a
bit awkward but we're not sure what to do. The addition of W-types and identity
is not a significant contribution by itself but we also don't want to just
handwave it. Maybe it could be turned into an appendix of this paper?

- exotic terms in HOAS

Exotic terms in CFTT are ruled out by the soundness & completeness of staging.
This is closely related to "adequacy" of HOAS in logical frameworks.

- The mixture of meta-language and object language can be very hard to read at times.

We agree that coloring could help and we could add it in a revision of the paper.

Also thanks for the other comments, we'll take them into account.

Review C
--------------------------------------------------------------------------------

Questions:

- In section 3.5, what becomes of the original example?

  The code output looks as if "modify' (λ x. <~x + 10>)"
  was inlined in the case branches:
  λ s. case b of
    True  → let s := 10; let s := s + 10; ((), s)
    False → let s := 20; let s := s + 10; ((), s)

  The fix would indeed have "join $ case b of ..."
  In a revision of the paper we could include these examples too.

- Extraction from Agda code

It wouldn't be hard! We could write a non-total interpreter for the object
language in Agda, and extract code for that, or just print Haskell code as
strings. We didn't implement these because we already had most of the examples
in executable form in Template Haskell.

- This example looks very much toyish and unmotivated

We don't think that it's toyish. In GHC it took significant engineering effort
and lots of tuning to reliably get similar code optimization for similar code
examples. There are lots of redexes and unfoldings in the non-optimized monadic
code, and GHC has to compute them using a general algorithm that also has to avoid
code size blow-ups.

- Rupicola: thanks, we will check this work out.

Addressing other questions/comments
------------------------------------------------------------

- "A weakness of the paper is that the intended model underlying the soundness
  of the approach is sketched very briefly."

We think that Section 2.4 is not very technically difficult when taken together with
Kovács22, for which it is basically an addendum; see also our answer to "Section
2.4" for Review B. We could spell out more details in an extended revision of
the paper or perhaps put them in an appendix.

- "MonadGen M": yes, we introduced things in a wrong order
- "reuse essentially all definitions from mtl": We'll write something clearer,
   here we meant mtl in a more language-agnostic way.

We'll act on the other comments as well.

### 2024-06-18 icfp24paper/paper.tex
fix

Modern functional programming supports many convenient abstractions. These often
come with significant runtime overheads. Sometimes the overheads are acceptable,
but in other cases compiler optimization is crucial. Monads in Haskell is an
example for the latter. Even the $\Reader$ monad, which is one of the simplest
in terms of implementation, yields large overheads when compiled without
optimizations. Consider the following:
\begin{alignat*}{3}
  &\vf :: \Int \to \Reader\,\Bool\,\Int \\
  &\vf\,\vx = \mdo\{\vb \fro \msf{ask}; \bif \vx\, \bthen \return (\vx + 10)\, \belse \return (\vx + 20)\}
\end{alignat*}
With optimizations enabled, GHC compiles this roughly to the code below:
\begin{alignat*}{3}
  &\vf :: \Int \to \Bool \to \Int \\
  &\vf = \lam \vx\,\vb.\, \bif \vb\, \bthen \vx + 10\, \belse \vx + 20
\end{alignat*}
Without optimizations we roughly get:
\begin{alignat*}{3}
  &\vf = \lam \vx.\,(\bind)\,\msf{monadReader}\,(\msf{ask}\,\msf{monadReaderReader})\,(\lam \vb.\,\bif \vb\\
  &\ind \bthen\return\,\msf{monadReader}\,(\vx + 10)\\
  &\ind \belse\hspace{0.25em}\return\,\msf{monadReader}\,(\vx + 20))
\end{alignat*}
Here, $\msf{monadReader}$ and $\msf{monadReaderReader}$ are runtime
dictionaries, respectively for the $\msf{Monad}$ and $\msf{MonadReader}$
instances, and, for example, $(\bind)\,\msf{monadReader}$ is a field projection
from the dictionary. This results from the dictionary-passing elaboration of
type classes \cite{DBLP:conf/popl/WadlerB89}. We get a runtime closure from the
$\lam \vb.\,...$ function, and $(\bind)$, $\msf{ask}$ and $\mbf{return}$ also
produce additional closures.

@@

In this appendix we describe the semantics of CFTT, from which the unstaging
algorithm is obtained, and also describe the soundness and completeness of the
algorithm. We borrow the setup from \cite{staged2ltt}; we use terminology and
definitions from there by default, and focus on differences and extensions.

>>>>

Modern functional programming supports many convenient abstractions. These often
come with significant runtime overheads. Sometimes the overheads are acceptable,
but in other cases compiler optimization is crucial. Monads in Haskell is an
example for the latter. Even the $\Reader$ monad, which is one of the simplest
in terms of implementation, yields large overheads when compiled without
optimizations. Consider the following:
\begin{alignat*}{3}
  &\vf :: \Int \to \Reader\,\Bool\,\Int \\
  &\vf\,\vx = \mdo\{\vb \fro \msf{ask}; \bif \vb\, \bthen \return (\vx + 10)\, \belse \return (\vx + 20)\}
\end{alignat*}
With optimizations enabled, GHC compiles this roughly to the code below:
\begin{alignat*}{3}
  &\vf :: \Int \to \Bool \to \Int \\
  &\vf = \lam \vx\,\vb.\, \bif \vb\, \bthen \vx + 10\, \belse \vx + 20
\end{alignat*}
Without optimizations we roughly get:
\begin{alignat*}{3}
  &\vf = \lam \vx.\,(\bind)\,\msf{monadReader}\,(\msf{ask}\,\msf{monadReaderReader})\,(\lam \vb.\,\bif \vb\\
  &\ind \bthen\return\,\msf{monadReader}\,(\vx + 10)\\
  &\ind \belse\hspace{0.25em}\return\,\msf{monadReader}\,(\vx + 20))
\end{alignat*}
Here, $\msf{monadReader}$ and $\msf{monadReaderReader}$ are runtime
dictionaries, respectively for the $\msf{Monad}$ and $\msf{MonadReader}$
instances, and, for example, $(\bind)\,\msf{monadReader}$ is a field projection
from the dictionary. This results from the dictionary-passing elaboration of
type classes \cite{DBLP:conf/popl/WadlerB89}. We get a runtime closure from the
$\lam \vb.\,...$ function, and $(\bind)$, $\msf{ask}$ and $\mbf{return}$ also
produce additional closures.

@@

In this appendix we describe the semantics of CFTT, from which the unstaging
algorithm is obtained, and also describe the soundness and stability of the
algorithm. We borrow the setup from \cite{staged2ltt}; we use terminology and
definitions from there by default, and focus on differences and extensions.
